---
title: "Ontario Cases - OSM Map"
author: "Chris Papalia"
date: "5/2/2020"
output: html_document
---

```{r setup, include=FALSE }
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

### Load the Libraries

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(maps)
library(sf)
library(mapproj)
library(OpenStreetMap)
library(ggrepel)
library(ggmap)
library(scales)
```


## Load and Clean the Data

Let's load the data from the website: [https://data.ontario.ca/dataset/confirmed-positive-cases-of-covid-19-in-ontario/resource/455fd63b-603d-4608-8216-7d8647f43350?view_id=95311e17-8314-42a7-82ae-05b55c839659](https://data.ontario.ca/dataset/confirmed-positive-cases-of-covid-19-in-ontario/resource/455fd63b-603d-4608-8216-7d8647f43350?view_id=95311e17-8314-42a7-82ae-05b55c839659), which allows us to download a list of confirmed cases.

I have already downloaded the files and put them in your working directory in Rstudio.cloud. Let's go through the cases as we look and inspect the file with the `glimpse()` command to get a sense of its structure. 
```{r}
ont_cases <- read_csv("ontario_covid_cases.csv")

ont_cases %>% 
glimpse()
```

We can also use the `summary()` command to get a sense of the data and describe some of the features we see. This gives a quick look at the basic elements and features of the data. For instance, we can see that when an variable is coded as a *character* datatype, it does not give us much insight in terms of grouping, so we'll need to make some changes to the data type of many of the variables.  

```{r}
ont_cases %>% 
  summary()
```


You can see that the structure has all character data types, so we should recode these character types as factors so that we can more easily manipulate the data. Fortunately, this data is already tidy because each row represents one observation (case) of COVID-19 and we won't need to do much in terms of tidying, but we'll need to do some basic data cleaning so that we have nicer data to work with. 


```{r}
ont_cases_clean <- 
ont_cases %>% 
  mutate_at(vars(Age_Group, Client_Gender, Reporting_PHU, Case_AcquisitionInfo, Outcome1), as.factor) %>% 
  # Change the selected vars to Factors to more easily work with them
  mutate(Age_Group = fct_recode(Age_Group, 
                                "0-20" = "<20",
                                "20-29" = "20s",
                                "30-39" = "30s",
                                "40-49" = "40s",
                                "50-59" = "50s",
                                "60-69" = "60s",
                                "70-79" = "70s",
                                "80-89" = "80s",
                                "90-99" = "90s", 
                                "NA" = "Unknown"
                                ) # Recode the Factors for Ages
         
         ) %>%   
  filter(Age_Group != "NA") %>%  # Remove the 7 Unknown Age_Group
  mutate(lat = Reporting_PHU_Latitude, 
         lon = Reporting_PHU_Longitude) %>%  # Create a Lat and Lon Variable
  mutate(Outcome = Outcome1, 
         Gender = Client_Gender) %>% 
  select(-`_id`,-Outcome1, -Client_Gender, -Reporting_PHU_Address:-Reporting_PHU_Longitude) # Remove the Variables not needed
```


## Explore the Data 

Now that we have cleaned the data, let's take a look at some basic elements of the data using some summaries and some plotting. Now that we made variables into factors, we are in good shape to examine the outputs in more detail.  

```{r}
ont_cases_clean %>% 
  summary()
```


We can go a little further in future anaysis to create some other grouping with the factors due to some small cases in some of the groups. The summary is a great place to start, though. So, let's take a look at distributions of age for these cases across Ontario. These plots will be quick and dirty. Hopefully the code is intutive also.

### Age Groups

I will walk us through the first part of the data to show how R thinks about individual observations of data. Then, I will use a different function that has this built in to it.

```{r}
ont_cases_clean %>% #Start with the Clean Data
  group_by(Age_Group) %>% #We want to group by the Age group of the case
  summarise(n = n()) %>% #Use the Summarize Function with the n() function to count how many people are in each group
  ggplot(aes(x = Age_Group, y = n))+ # Create a plot later that has Age on x and count on y
  geom_col() # Make this a column plot
```


The same plot a different way. The geom_bar() function uses the stat = count() at its core, so we don't need to create a summary before plotting. This is useful when your data is already tidy. 

```{r}
ont_cases_clean %>% 
  ggplot(aes(x = Age_Group))+
geom_bar()
```


Let's see if we can find some relationships between some of the variables. Let's see if the outcome varies much with the age groups. We can look at this by count and by percent. 

```{r}
ont_cases_clean %>% 
  ggplot(aes(x = Age_Group, fill = Outcome))+
  geom_bar()
```


```{r}
ont_cases_clean %>% 
  ggplot(aes(x = Age_Group, fill = Outcome))+
  geom_bar(aes(y = (..count..)/sum(..count..)))+
  scale_y_continuous(labels = percent_format())+
  labs(x = "Age Group", 
       y = "Percent of Cases",
       title = "The Number of COVID-19 Cases by Age Group",
       subtitle = "Colour Indicates Outcome")+
  theme_light()
  
```


### Gender

Firstly, with gender, we will need to remove the 2 others, the 2 transgender, and the 149 unknowns. Then, we'll look at the data in a side-by-side bar chart.


```{r}
ont_cases_clean %>%
  filter(Gender %in% c("MALE", "FEMALE")) %>% 
  ggplot(aes(x = Age_Group, fill = Gender))+
  geom_bar(position = "dodge")+
  theme_minimal()+
  labs(x = "Age Group",
       y = "Count", 
       title = "The Number of COVID-19 Cases in Ontario by Gender", 
       subtitle = "Colour Indicates Gender")
```

Let's wonder, what does this reveal about the confirmed cases in Ontario? 


I wonder if maybe  there is a relationship between whether cases have been fatal more often in men or women bases on the data. Note, we're only looking at cases that are have been resolved in one way or another. 

```{r}
ont_cases_clean %>% 
  filter(Gender %in% c("MALE", "FEMALE")) %>% 
  filter(Outcome != "Not Resolved") %>% 
  ggplot(aes(x = Gender, fill = Outcome))+
  geom_bar(position = "fill")+
  scale_y_continuous(labels = percent_format())

```



## Mapping Locations for Cases Across Southern Ontario

Something that I haven't tried to yet is create a map using the R software, so it was exciting to create a map using the Open Street Maps package. Unforunatey, only some countries have maps built into the regular r packages, and I couldn't find a map of Ontario very easily. 

To start, let's look at a basis map of the lat and lon coordinates in a scatter plot. In order to do that, we need to summarize the data by the lat and lon coordinates of the reporting Ontario Health Units. This new data will count all the cases processed at the health unit and keep the lat and lon coordinates of each of the health units for a map. 

```{r}
ont_cases_loc <- 
ont_cases_clean %>% # Start with our cleaned data
  group_by(Reporting_PHU) %>%  # Group each case by the Health Unit
  summarise(cases = n(), # Count How many cases at each unit using n()
            lat = mean(lat), # create a variable lat (note that I needed to use mean because it aggregates all the data)
            lon = mean(lon) # create a variable lon (note that I needed to use mean because it aggregates all the data)
            ) %>% 
  ungroup() # Ungroups the data to make it a data class rather than grouped data

ont_cases_loc %>% 
  arrange(desc(cases))
```


### Map these locations on a map of Canada

The only map that I could find built into R was a map of Canada. This would be helpful if I could filter it by province, but alas, I can't get any more detailed than regions = Canada. 


Essentialy what I need to do is map a scatterplot onto a map. Here's a scatterplot function using geom_point() and the geom_sf() function is used to keep the map coordinates in perspective.

```{r}
ggplot()+
  geom_point(data = ont_cases_loc, aes(x = lon, y = lat, size = cases))+
  geom_sf()
```

Here is the output of a map of Canada.

```{r}
ont_cases_loc %>% 
  ggplot(aes(x = lon, y = lat, size = cases))+
  geom_point()+
  coord_map()+
  borders(database = "world", regions = "canada")+
  theme_minimal()+
  scale_size(name = "Cases", breaks = c(100, 500, 1000, 5000), range = c(1, 5))+
  labs(x = "", y = "")
```


### Map these locations on a street map

We will create an object that pulls from the Open Street Map database to plot the points on. We will call the object OSM_South_Ont to represent Southern Ontario for our plot. 

```{r}
OSM_South_Ont <- openmap(c(41.73, -83.84), c(45.6, -73.85), type = "esri-topo")
```

Once, we assign the map to an object, we can use the `ggplot()` function to plot it. As with any `ggplot()`, we can add layers onto the plot to create a map of the amount of cases that are seen at each Provincial Health Unit. 

Here is the layer that we will build our plot on.
```{r}
autoplot(OSM_South_Ont)
```

One of the drawbacks of the OSM Map is that the latitude and longitude coordinates are enlarged in their scales, so the actual latitude and longitude coordinates do not match the map. This is a drawback of the OSM Maps, but it's possible to convert the actual latitude and longitude into coordinates that match up with the OSM Map. 

Let's work through that by scaling the coordinates in `ont_cases_loc` data while preserving the Reporting Health Units. We can scale these coordinates using the `projectMercator()` function in the OpenStreetMaps package. 


```{r}
osm_plot <- as.data.frame(projectMercator(ont_cases_loc$lat, ont_cases_loc$lon)) #this takes the lat and lon coordinates as converted and puts them into a data frame to use to plot. We will then add onto it.

osm_plot %>% 
  head(3)
```

By inspecting this, we see the function has created an x and y coordinate for the latitude and longitude inputs we gave them. They're also in order of the Reporting Health Units. This is impotant so that we can preserve the proper Health Unit with the correct coordinates. 

Now, we'll add new variables using the mutate function (transmute) to create the other elements of the dataframe we're going to plot and remove the old x and y coordinates. The transmute function removes the variables we don't specify. 

```{r}
osm_plot %>% 
  transmute(lat = x, 
         lon = y, 
         health_unit = ont_cases_loc$Reporting_PHU,
         cases = ont_cases_loc$cases) -> osm_plot_RHU

osm_plot_RHU %>% 
  head(3)
```



Now that we have the Reporting Heath Units with coordinates that scale to the map, we are ready to make our plot.

First, though, we need to remove some of the Reporting Health Units that are outside of our Southern Ontario map boundaries. (Sorry for those from this area!). We'll create a dataframe that filters those Health Units that are outside the boundaries of the plot. 

```{r}
osm_plot_RHU_South <- 
osm_plot_RHU %>% 
  filter(lon < 5700000) %>% 
  filter(lat < -8400000)
```


```{r, fig.width=7, fig.height=5}
autoplot(OSM_South_Ont)+
  geom_point(data = osm_plot_RHU_South, aes(x = lat, y = lon, size = cases, colour = health_unit), alpha = .8)+
  scale_x_continuous(breaks = NULL)+
  scale_y_continuous(breaks = NULL)+
  scale_color_discrete(guide = FALSE )+
  scale_size(name = "Cases", breaks = c(100, 500, 1000, 5000), range = c(1, 10))+
  labs(x = "", 
       y = "", 
       title = "The Number of Confirmed Cases of COVID-19 Reported in Southern Ontario",
       subtitle = "Grouped by Ontario Health Units")+
  geom_text_repel(data = osm_plot_RHU_South, aes(x = lat, y = lon, label = health_unit), size = 2)+
  theme_minimal()+
  theme(legend.justification = c(1,0), legend.position = c(.95,.1))

```


